Stateful Forward Reasoning

### 2025-01-31 paper.tex
Reformat; semantic linebreaks; some typos

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

An important contextualization is that the forward reasoning procedure in this paper takes place
within aesop's routine procedure. We will focus on examples that highlight this style of proof
while noting that in practice, a combination of both forwards and backwards reasoning are used 
\section{Preliminaries}

@@

\paragraph{Forward Rules.}
Consider the following lemma :
\begin{lem}[\lean{le_trans}]
  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds :
  $$a \leq b \rightarrow b \leq c \rightarrow a \leq c$$
\end{lem}

@@

A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in
the context.
It must satisfy the following conditions:
\begin{itemize}
\item For each slot $i$ of $r$, if $M(i)$ is defined, then
    $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
    The domain of this substitution is a subset of the non-maximal input hypotheses.
\item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined,
    $\sigma_i$ and $\sigma_j$ are compatible.
\item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of
    $r$, $M(j)$ is also defined.

\end{itemize}
A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and
\textit{incomplete} otherwise.
The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ 
is defined.
The \textit{substitution of a match} $M$, $sub(M)$, is the union of the
substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
This union is defined since the $\sigma_i$ are pairwise compatible.
This means that $sub(M)$ contains only one assignement per variable.
The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the
shared variables of $lvl(M)$.

@@

We then aim to construct matches iteratively : 
We consider each hypotheses one by one, assign them to their
respective slots and check whether there exist compatible matches
which they can extend.
The comptibility check means that all the shared variables of the considered 
hypotheses need to be compatible.
It would be costly to do this comparison everytime and indeed
we choose instead to maintain what we call \textit{variable maps} whose purpose
is to save the current partial matches.
Each variable map is associated to one shared variable, these are the
variables dictate compatibility.

@@

Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$,
the map sending the variable of $T_i$ to their specific instantiation for $h$.
The compatibility check that looks for suitable partial matches is simply computed by
$$\bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right).$$

Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant 
variable in $m$ then we can check for compatible hypotheses that would extend $m$
by computing 
$$\bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right).$$

@@

At the beginning, for the rule \textit{le\_trans}, we notice that the
only shared variable is ${y}$. Now the hypothesis $(h_2)$ unifies both
with slot 1 $x \leq y$ and slot 2 $y \leq z$.
This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and 
$\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$

@@

\paragraph{Example 2.}
We now consider another example with more than one shared variable.
Let $T_1, T_2$ be two proposition that both depend on the variable $x$ and $y$.
For simplicity, the rule considred will be $T_1(x,y) \to T_2(x,y) \to R$.
The hypotheses added to the context will be, in order,
$h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.

@@

  As this hypothesis is in the first slot, we create the corresponding level 1 partial
  match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could
  extend the match.
  Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables
  appearing in the slots $i < 2$ with the variables appearing in slot $2$.
  Thus $s(2) = \{a,b\}$ and we compute :
  $$\bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right)
    = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset $$
  Since the set is empty, the match can not be currently extended.
  We turn to the remaining hypothesis in the queue and add it to the maps.

@@

  We are now ready to look for potential matches that could be extended by $h_3$.
  In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{a,b\}$.
  Then,
  $$\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right)
    = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\} $$.

@@

\xcom{Here are some notes about the implementation.}
The index used to match hypotheses with their matching rules are implemented as
discrimination trees. A discrimination tree is a trie-like data
structure that maps expressions to arbitrary data.
In \cite{}, it is shown to be good solution for the problem of finding
terms (here the types of hypotheses) that might unify with other terms
(here the types of the input hypotheses).
In Lean 4, discrimination trees are also used to index typeclass
instances and simplifier lemmas.

\xcom{Incrementality?}
In lean, the type of hypotheses can sometimes change while the hypothesis'
identifier remains the same. This means that one needs to track the types
of the hypotheses as well. We call this tracking \textit{goal diffs}, they
highlight the differences between old and updated goal states.
The reference to hypotheses which changing types get deleted from the maps
and the slot. We then add them again as a new hypothesis.

The maps described in the last section are all implemented as persistent
hash maps \xcom{Why did we not use discrimination trees? $+$
(pers. important cause we store related forward state for many goal.) }.

One of the more costly operation that we need to run multiple times is the
unification \xcom{defeq?} between the hypotheses and the types of the
input hypotheses associated to rules.
We implement a variant of Lean's unification process as follows :
The idea is that we wish to normalize the types as much as possible

@@

Analysing the both algorithms of both the stateless and stateful approach, we 
can draw the following conclusions about their relative performance :

\begin{itemize}
    \item The stateful algorithm maybe be expected to perform poorly when the
    context contains many hypotheses that almost complete rules.
    \item The stateless algorithm should perform decently well on average and 
    especially when only the last premises are absent; so the rules get never
    selected by the stateless index
\end{itemize}

@@

We present here a number of results obtained by running synthetic tests.
By synthetic we mean that these tests we're constructed solely for the
purpose of demonstrating the different complexity behavior of both
algorithms by scaling some parameter. (e.g. the number of rules.)
The results of the tests are an average of multiple runs.

In \ref{fig:erase}, the step is set up as follows: 
Given some rule $r$ with $6$ \xcom{motivate 6} premises, 
we run the procedure in a context with 5 of the 6 premises present.
The numbers on the $x$ axis represent which hypothesis is missing
in the order of consideration by the algorithm.
In a sense, this is how \textit{deep} the algorithm investigate before
it stops.
At $0$, the rule are not selected by the index so the procedure stops.
The stateless algorithm gets progressively slower the further the removed
premises is located.

>>>>

\begin{abstract}
  TODO

@@

An important contextualization is that the forward reasoning procedure in this paper takes place within Aesop's routine procedure.
We will focus on examples that highlight this style of proof while noting that in practice, a combination of both forwards and backwards reasoning are used.

\section{Preliminaries}

@@

\paragraph{Forward Rules.}
Consider the following lemma:
\begin{lem}[\lean{le_trans}]
  Given a preorder on some type $\alpha$ and elements $a,b,c$ of type $\alpha$, the following holds:
  $$a \leq b \rightarrow b \leq c \rightarrow a \leq c$$
\end{lem}

@@

A \textit{match} $M$ for $r$ is a partial map from slots of $r$ to hypotheses in the context.
It must satisfy the following conditions:
\begin{itemize}
  \item For each slot $i$ of $r$, if $M(i)$ is defined, then $M(i) \equiv_{[\sigma_i]} T_i$ with a substitution $\sigma_i$.
        The domain of this substitution is a subset of the non-maximal input hypotheses.
  \item For all slots $i$ and $j$ of $r$ such that $M(i)$ and $M(j)$ are defined, $\sigma_i$ and $\sigma_j$ are compatible.
  \item $M$ is downward-closed: if $M(i)$ is defined, then for any slot $j < i$ of $r$, $M(j)$ is also defined.
\end{itemize}

A match of $r$ is \textit{complete} if $M(i)$ is defined for every slot $i$ of $r$, and \textit{incomplete} otherwise.
The \textit{level of a match} $M$, $lvl(M)$, is the maximal $i$ such that $M(i)$ is defined.
The \textit{substitution of a match} $M$, $sub(M)$, is the union of the substitutions $\sigma_i$ for each $i$ such that $M_i$ is defined.
This union is defined since the $\sigma_i$ are pairwise compatible.
This means that $sub(M)$ contains only one assignement per variable.
The set of \textit{shared variables of a match} $M$ is $s(lvl(M))$ the shared variables of $lvl(M)$.

@@

We then aim to construct matches iteratively: we consider each hypotheses one by one, assign them to their respective slots and check whether there exist compatible matches which they can extend.
The compatibility check means that all the shared variables of the considered hypotheses need to be compatible.
It would be costly to do this comparison every time and indeed we choose instead to maintain what we call \textit{variable maps} whose purpose is to save the current partial matches.
Each variable map is associated to one shared variable, these are the variables that dictate compatibility.

@@

Given a hypothesis $h : T$ that matches $T_i$ and $\sigma_h$, the map sending the variable of $T_i$ to their specific instantiation for $h$.
The compatibility check that looks for suitable partial matches is simply computed by
\[
  \bigcap_{s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)
\]

Similarly, given a match $m$ and $\sigma_m$ the substitution map of the relevant  variable in $m$ then we can check for compatible hypotheses that would extend $m$ by computing
\[
  \bigcap_{s(lvl(M) + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)
\]

@@

At the beginning, for the rule \textit{le\_trans}, we notice that the only shared variable is ${y}$.
Now the hypothesis $(h_2)$ unifies both with slot 1 $x \leq y$ and slot 2 $y \leq z$.
This updates the map $\eta_y(y \mapsto 302, 1)$ with $\{h_2\}$ and $\eta_y(y \mapsto 301, 2)$ with $\{h_2\}$.

@@

\paragraph{Example 2.}
We now consider another example with more than one shared variable.
Let $T_1, T_2$ be two proposition that both depend on the variable $x$ and $y$.
For simplicity, the rule considered will be $T_1(x,y) \to T_2(x,y) \to R$.
The hypotheses added to the context will be, in order, $h_1 : T_2(a,a)$, $h_2 : T_1(a,b)$ and $h_3 : T_2(a,b)$.

@@

As this hypothesis is in the first slot, we create the corresponding level 1 partial match $m_0 = {(h_2 : T_1(a, b))}$ for which we check for suitable hypotheses that could extend the match.
Note that  $s(lvl(m_0) + 1) = s(2)$ which is the intersection of the variables appearing in the slots $i < 2$ with the variables appearing in slot $2$.
Thus $s(2) = \{a,b\}$ and we compute:
\[
  \bigcap_{x\in s(lvl(m_0) + 1)} \eta_x \left(i + 1, \sigma_{m_0}(x)\right) = \eta_x(2, a) \cap \eta_y(2, b) = \emptyset
\]
Since the set is empty, the match can not be currently extended.
We turn to the remaining hypothesis in the queue and add it to the maps.

@@

We are now ready to look for potential matches that could be extended by $h_3$.
In this particuliar case, since $h_3$ is in slot $2$ we look for $s(2) = \{a,b\}$.
Then,
\[
\bigcap_{x\in s(2)} \mu_x \left(i - 1, \sigma_h(x)\right) = \mu_x(1, a) \cap \mu_y(1, b) = \{m_0\}
\]

@@

\xcom{Here are some notes about the implementation.}
The index used to match hypotheses with their matching rules are implemented as discrimination trees.
A discrimination tree is a trie-like data structure that maps expressions to arbitrary data.
In~\cite{TODO}, it is shown to be good solution for the problem of finding terms (here the types of hypotheses) that might unify with other terms (here the types of the input hypotheses).
In Lean 4, discrimination trees are also used to index typeclass instances and simplifier lemmas.

\xcom{Incrementality?}
In Lean, the type of hypotheses can sometimes change while the hypothesis' identifier remains the same.
This means that one needs to track the types of the hypotheses as well.
We call this tracking \textit{goal diffs}, they highlight the differences between old and updated goal states.
The reference to hypotheses which changing types get deleted from the maps and the slot.
We then add them again as a new hypothesis.

The maps described in the last section are all implemented as persistent hash maps.
\xcom{Why did we not use discrimination trees? $+$ (pers. important cause we store related forward state for many goal.)}

One of the more costly operation that we need to run multiple times is the unification \xcom{defeq?} between the hypotheses and the types of the input hypotheses associated to rules.
We implement a variant of Lean's unification process as follows: the idea is that we wish to normalize the types as much as possible.

@@

Comparing the stateless and stateful approach, we can draw the following conclusions about their relative performance:

\begin{itemize}
    \item The stateful algorithm maybe be expected to perform poorly when the context contains many hypotheses that almost complete rules.
    \item The stateless algorithm should perform decently well on average and especially when only the last premises are absent, so the rules get never selected by the stateless index.
\end{itemize}

@@

We present here a number of results obtained by running synthetic tests.
By synthetic we mean that these tests we're constructed solely for the purpose of demonstrating the different complexity behavior of both algorithms by scaling some parameter (e.g.\ the number of rules).
The results of the tests are an average of multiple runs.

In Fig.~\ref{fig:erase}, the step is set up as follows: given some rule $r$ with $6$ \xcom{motivate 6} premises, we run the procedure in a context with 5 of the 6 premises present.
The numbers on the $x$ axis represent which hypothesis is missing in the order of consideration by the algorithm.
In a sense, this is how \textit{deep} the algorithm investigate before it stops.
At $0$, the rule are not selected by the index so the procedure stops.
The stateless algorithm gets progressively slower the further the removed premises is located.
